---
title: "Benchmarking alternative implementations for a simple rowwise job"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking alternative implementations for a simple rowwise job}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Vignette overview

The goal of this vignette is to compare the performances of `lay()` to alternatives.
As you will see, the code using `lay()` is both simple and efficient.
The only alternative that is clearly more efficient is the one labelled below "*vectorized*".
Unfortunately, such a vectorized approach imply to refer explicitly to all column names which data are used.
Furthermore, such a vectorized approach is not applicable generally, as it can only deal with logical
and arithmetic operators and does allow the use of other types of functions.


## Loading packages

This vignette requires you to load the following packages:

```{r load_pkg, message=FALSE}
library(dplyr)      ## for many things
library(lay)        ## for lay() and the data
library(tidyr)      ## for pivot_longer() and pivot_wider()
library(purrr)      ## for pmap_lgl()
library(data.table) ## for an alternative to base and dplyr
library(bench)      ## for running the benchmark
library(forcats)    ## for sorting levels in benchmark plot with fct_reorder()
```

Please install them if they are not present on your system.


## An example of a rowwise task

Consider the dataset `drugs_full` from our package {lay}:
```{r drugs_full}
drugs_full
```

In this dataset, all columns but `caseid` record the use of pain relievers for non medical purpose.

For each drug there is a certain number of users and non-users:
```{r drugs_full_summary}
drugs_full %>%
  pivot_longer(-caseid, names_to = "drug", values_to = "used") %>%
  count(drug, used) %>%
  mutate(used = if_else(used == 1, "have_used", "have_not_used")) %>%
  pivot_wider(names_from = used, values_from = n)
```

In this vignette, we will compare the efficiency of alternative ways to create a new variable named `everused` which will indicate if each respondent as used any of the considered pain relievers for non medical purpose or not.

## Benchmark

Let's compare the running time of different methods to do this job:

```{r bench_run1}
drugs_full_dt <- data.table(drugs_full) ## coercion to data.table

benchmark1 <- mark(
  vectorized = {
    drugs_full %>%
      mutate(everused = codeine | hydrocd | methdon | morphin | oxycodp | tramadl | vicolor)},
  lay = {
    drugs_full %>%
      select(-caseid) %>%
      mutate(everused = lay(., any))},
  lay_alternative = {
    drugs_full %>%
      mutate(everused = lay(across(-caseid), any, .method = "tidy"))},
  c_across = {
    drugs_full %>%
      rowwise() %>%
      mutate(everused = any(c_across(-caseid))) %>%
      ungroup()},
  pivot_pivot = {
    drugs_full %>%
      pivot_longer(-caseid) %>%
      group_by(caseid) %>%
      mutate(everused = any(value)) %>%
      ungroup() %>%
      pivot_wider()},
  pmap = {
    drugs_full %>%
      mutate(everused = pmap_lgl(across(-caseid), ~ any(...)))},
  data.table = {
    drugs_full_dt[, ..I := .I]
    drugs_full_dt[, everused := any(.SD), by = ..I, .SDcols = -"caseid"]},
  apply = {
    drugs_full %>%
      mutate(everused = apply(across(-caseid), 1, any))},
  'for' = {
    everused <- logical(nrow(drugs_full))
    columns_in <- colnames(drugs_full) != "caseid"
    for (i in seq_len(nrow(drugs_full))) everused[i] <- any(drugs_full[i, columns_in])},
  iterations = 5,
  time_unit = "ms",
  check = FALSE
  )
```

Here are the results of this first benchmark:
```{r bench_results1}
benchmark1 %>%
  arrange(median)
```

```{r bench_plot1, fig.width=7}
benchmark1 %>%
  mutate(expression = fct_reorder(as.character(expression), median, .desc = TRUE)) %>%
  plot()
```

Note that the scale of the plot is logarithmic.

## Benchmark on smaller dataset

Let's repeat our benchmark using a smaller dataset:

```{r bench_run2}
drugs_dt <- data.table(drugs) ## coercion to data.table

benchmark2 <- mark(
  vectorized = {
    drugs %>%
      mutate(everused = codeine | hydrocd | methdon | morphin | oxycodp | tramadl | vicolor)},
  lay = {
    drugs %>%
      select(-caseid) %>%
      mutate(everused = lay(., any))},
  lay_alternative = {
    drugs %>%
      mutate(everused = lay(across(-caseid), any, .method = "tidy"))},
  c_across = {
    drugs %>%
      rowwise() %>%
      mutate(everused = any(c_across(-caseid))) %>%
      ungroup()},
  pivot_pivot = {
    drugs %>%
      pivot_longer(-caseid) %>%
      group_by(caseid) %>%
      mutate(everused = any(value)) %>%
      ungroup() %>%
      pivot_wider()},
  pmap = {
    drugs %>%
      mutate(everused = pmap_lgl(across(-caseid), ~ any(...)))},
  data.table = {
    drugs_dt[, ..I := .I]
    drugs_dt[, everused := any(.SD), by = ..I, .SDcols = -"caseid"]},
  apply = {
    drugs %>%
      mutate(everused = apply(across(-caseid), 1, any))},
  'for' = {
    everused <- logical(nrow(drugs))
    columns_in <- colnames(drugs) != "caseid"
    for (i in seq_len(nrow(drugs))) everused[i] <- any(drugs[i, columns_in])},
  iterations = 30,
  time_unit = "ms",
  check = FALSE
  )
```

Here are the results of this second benchmark:
```{r bench_results2}
benchmark2 %>%
  arrange(median)
```

```{r bench_plot2, fig.width=7}
benchmark2 %>%
  mutate(expression = fct_reorder(as.character(expression), median, .desc = TRUE)) %>%
  plot(type = "violin")
```

Note that the scale of the plot is logarithmic.

## Benchmark environment

```{r session}
sessionInfo()
```

