---
title: "Benchmarks"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Article overview

The goal of this Article is to compare the performances of `lay()` to alternatives described [here](https://courtiol.github.io/lay/articles/alternatives.html).
As you will see, the code using `lay()` is quite efficient.
The only alternative that is clearly more efficient is the one labeled below "*vectorized*".
Unfortunately, such a vectorized approach imply to refer explicitly to all column names which data are used.
Furthermore, such a vectorized approach is not applicable generally, as it can only deal with logical
and arithmetic operators and does allow the use of other types of functions.


## Loading packages

This Article requires you to load the following packages:

```{r load_pkg, message=FALSE}
library(lay)        ## for lay() and the data
library(dplyr)      ## for many things
library(tidyr)      ## for pivot_longer() and pivot_wider()
library(purrr)      ## for pmap_lgl()
library(slider)     ## for slide()
library(data.table) ## for an alternative to base and dplyr
library(bench)      ## for running the benchmarks
library(forcats)    ## for sorting levels in plot with fct_reorder()
```

Please install them if they are not present on your system.


## An example of a rowwise task

Consider the dataset `drugs_full` from our package {lay}:
```{r drugs_full}
drugs_full
```

In this dataset, all columns but `caseid` record the use of pain relievers for non medical purpose.

For each drug there is a certain number of users and non-users:
```{r drugs_full_summary}
drugs_full |>
  pivot_longer(-caseid, names_to = "drug", values_to = "used") |>
  count(drug, used) |>
  mutate(used = if_else(used == 1, "have_used", "have_not_used")) |>
  pivot_wider(names_from = used, values_from = n)
```

In this Article, we compare the efficiency of alternative ways to create a new variable named `everused` which indicates if each respondent has used any of the considered pain relievers for non medical purpose or not.

We will run benchmarks on the dataset `drugs_full` and its `r nrow(drugs_full)` rows, as well as on a subset of this data called `drugs` that only contains `r nrow(drugs)` rows.

## Benchmarks on the full dataset (`r nrow(drugs_full)` rows)

Let's compare the running time of different methods to do this job on the full dataset:

```{r bench_run1}
drugs_full_dt <- data.table(drugs_full) ## coercion to data.table

benchmark1 <- mark(
  vectorized = {
    drugs_full |>
      mutate(everused = codeine | hydrocd | methdon | morphin | oxycodp | tramadl | vicolor)},
  lay = {
    drugs_full |>
      select(-caseid) |>
      mutate(everused = lay(pick(everything()), any))},
  lay_alternative = {
    drugs_full |>
      mutate(everused = lay(pick(-caseid), any, .method = "tidy"))},
  c_across = {
    drugs_full |>
      rowwise() |>
      mutate(everused = any(c_across(-caseid))) |>
      ungroup()},
  pivot_pivot = {
    drugs_full |>
      pivot_longer(-caseid) |>
      group_by(caseid) |>
      mutate(everused = any(value)) |>
      ungroup() |>
      pivot_wider()},
  pmap = {
    drugs_full |>
      mutate(everused = pmap_lgl(pick(-caseid), ~ any(...)))},
  slider = {
    drugs_full |>
      mutate(everused = slide_vec(pick(-caseid), any))},
  data.table = {
    drugs_full_dt[, ..I := .I]
    drugs_full_dt[, everused := any(.SD), by = ..I, .SDcols = -"caseid"]},
  apply = {
    drugs_full |>
      mutate(everused = apply(pick(-caseid), 1, any))},
  'for' = {
    everused <- logical(nrow(drugs_full))
    columns_in <- colnames(drugs_full) != "caseid"
    for (i in seq_len(nrow(drugs_full))) everused[i] <- any(drugs_full[i, columns_in])},
  iterations = 5,
  time_unit = "ms",
  check = FALSE
  )
```

Here are the results of this first series of benchmarks:
```{r bench_results1}
benchmark1 |>
  arrange(median)
```

```{r bench_plot1, fig.width=7}
benchmark1 |>
  mutate(expression = fct_reorder(as.character(expression), median, .desc = TRUE)) |>
  plot()
```

Note that the x-axis of the plot is on a logarithmic scale.

## Benchmarks on a subset of the data (`r nrow(drugs)` rows)

Let's repeat our benchmarks using a only a subset of the original dataset:

```{r bench_run2}
drugs_dt <- data.table(drugs) ## coercion to data.table

benchmark2 <- mark(
  vectorized = {
    drugs |>
      mutate(everused = codeine | hydrocd | methdon | morphin | oxycodp | tramadl | vicolor)},
  lay = {
    drugs |>
      select(-caseid) |>
      mutate(everused = lay(pick(everything()), any))},
  lay_alternative = {
    drugs |>
      mutate(everused = lay(pick(-caseid), any, .method = "tidy"))},
  c_across = {
    drugs |>
      rowwise() |>
      mutate(everused = any(c_across(-caseid))) |>
      ungroup()},
  pivot_pivot = {
    drugs |>
      pivot_longer(-caseid) |>
      group_by(caseid) |>
      mutate(everused = any(value)) |>
      ungroup() |>
      pivot_wider()},
  pmap = {
    drugs |>
      mutate(everused = pmap_lgl(pick(-caseid), ~ any(...)))},
  slider = {
    drugs |>
      mutate(everused = slide_vec(pick(-caseid), any))},
  data.table = {
    drugs_dt[, ..I := .I]
    drugs_dt[, everused := any(.SD), by = ..I, .SDcols = -"caseid"]},
  apply = {
    drugs |>
      mutate(everused = apply(pick(-caseid), 1, any))},
  'for' = {
    everused <- logical(nrow(drugs))
    columns_in <- colnames(drugs) != "caseid"
    for (i in seq_len(nrow(drugs))) everused[i] <- any(drugs[i, columns_in])},
  iterations = 30,
  time_unit = "ms",
  check = FALSE
  )
```

Here are the results of this second series of benchmarks:
```{r bench_results2}
benchmark2 |>
  arrange(median)
```

```{r bench_plot2, fig.width=7}
benchmark2 |>
  mutate(expression = fct_reorder(as.character(expression), median, .desc = TRUE)) |>
  plot(type = "violin")
```

Note again that the x-axis of the plot is on a logarithmic scale.

## Benchmarks' environment

```{r session}
sessionInfo()
```

